Viscous Erosion in Porous Media

### 2018-02-22 docs/methods/erosion_methods.tex
NM
Making minor changes to intro

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction\label{s:intro}}
Erosion is a fluid-mechanical process that is responsible for a variety
of phenomena in geophysical fluids and other processes.  For example,
wind erosion causes sand dunes to be  longer on the windward
side~\cite{han1969}, and erosion is responsible for branching structures
of streams and rivers~\cite{coh-dev-sey-yi-szy-rot2015}.  Erosion is
also important in the biophysical community. Plaque erosion in arteries
is key to understanding heart disease~\cite{sha2002,
gro-gij-van-fer-hat-van-yua-wen2007}, and the separation of biofilms can
be explained by erosion~\cite{pic-van-hei2000}.  Another application,
which motivates the current work, is erosion caused by groundwater flow.
Here, a slow moving fluid erodes channels in the ground, and this
affects the fluid transport.  In all these examples, there is a feedback
process between the flow and the eroding bodies---the shape of the
bodies affect the flow, while the flow changes the shape by fluidizing
each body.  This fluid-mechanical feedback system has been simulated in
idealized cases such as a single eroding two-dimensional body in a
uniform flow~\cite{moo-ris-chi-zha-she2013} or a single mobile eroding
three-dimensional body~\cite{mit-spa2016}. However, to better understand
the effect of erosion on groundwater flow, multiply-connected domains
must be considered.  Here we simulate two-dimensional erosion, in
multiply-connected domain such as the porous geometry of many
groundwater flows (Fig.~\ref{fig:50bodies}).

Because of the length and velocity scales of groundwater flow, we
restrict our attention to small Reynolds number flows.  For groundwater
flow, Darcy's law is often used, but this does not resolve the pore
structure and assumes a homogeneous geometry which is often not valid.
Instead, we use the incompressible Stokes equations to represent the
flow, and the individual pores are resolved which results in a complex
multiply-connected geometry.  A boundary integral equation is
well-suited for solving the incompressible Stokes equations in a complex
geometry since it can resolve individual grains and the non-negligible
interactions between bodies are resolved. Moreover, high-order accuracy
and computational efficiency are obtained. To link the eroding body and
fluid dynamics, the body is eroded in the normal direction at a rate
that is proportional to the magnitude of the shear
stress~\cite{ris-moo-chi-she-zha2012}.  Tracking the interface of
fluid-structure interaction problems with a long time horizon is
particularly---the mesh is often distorted and tangled by the dynamics
and regions of high curvature develop.  These two issues are resolved
by, first, introducing a tangential velocity field to maintain a grid
that is equispaced in arclength, but that does not affect the dynamics,
and, second, a curvature penalization term.  Furthermore, the frame of
reference is changed to \thL~variables so that the curvature term
appears as a linear diffusive term that can easily be treated implicitly
to avoid stiffness.

@@

At the second stage of RK2, we discretize \eqref{eq:mu}--\eqref{eqn:thetaevo3} as 
\begin{align}
\label{eq:mu1d}
& \mu_k^{(n+1)} = \mu_k^{(n)} \exp \left( \eps k^2 {\Dt} \, \zeta^{(n+1/2)} \right) \\
\label{eq:L1d}
& \frac{L^{(n+1)} - L^{(n)}}{\Dt} = \Mterm^{(n+1/2)}, \\
\label{eq:th1d}
& \frac{ \mu_k^{(n+1)} \thhat_k^{(n+1)} - \mu_k^{(n)} \thhat_k^{(n)}}{\Dt} =
\mu_k^{(n+1/2)} \hat{\NLterm}_k^{(n+1/2)}.
\end{align}
Once again, we must eventually compute the geometry at step $n+1$, and so we use \eqref{eq:mu1d} to solve \eqref{eq:th1d} for $\theta^{(n+1)}$, giving
\begin{align}
\thhat_k^{(n+1)} = & \thhat_k^{(n)} \exp \left( - \epsilon k^2 \Dt \elfun^{(n+1/2)} \right) + \Dt \, \hat{\NLterm}_k^{(n+1/2)} \exp \left( - \frac{1}{2} \eps k^2 \Dt \left( 2 \elfun^{(n+1/2)} - \elfun^{(n)} \right) \right).
\end{align}
Again, both exponential terms act to reduce high-frequency oscillations in shape.

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction\label{s:intro}}
Erosion is a fluid-mechanical process responsible for a variety
of phenomena in geophysical and other settings.  For example,
wind erosion causes sand dunes to be  longer on the windward
side~\cite{han1969}, while stream-bed erosion is responsible for branching structures
of rivers~\cite{coh-dev-sey-yi-szy-rot2015}.  Erosion is
also important in biophysical applications, for example plaque erosion in arteries
is key to understanding heart disease~\cite{sha2002,
gro-gij-van-fer-hat-van-yua-wen2007}, and the separation of biofilms can
be explained by erosion~\cite{pic-van-hei2000}.  Another application,
which motivates the current work, is erosion caused by groundwater flow.
Here, a slow moving fluid erodes channels in the ground, and this
affects the fluid transport.  In all these examples, there is a feedback
process between the flow and the eroding bodies---the shape of the
bodies affect the flow, while the flow changes the shape by fluidizing
each body.  This fluid-mechanical feedback system has been simulated in
idealized cases such as a single eroding two-dimensional body in a
uniform flow~\cite{moo-ris-chi-zha-she2013} or a single mobile eroding
three-dimensional body~\cite{mit-spa2016}. However, to better understand
the effect of erosion on groundwater flow, multiply-connected domains
must be considered.  Here we simulate two-dimensional erosion, in
multiply-connected domain such as the porous geometry of many
groundwater flows (Fig.~\ref{fig:50bodies}).

Because of the length and velocity scales of groundwater flow, we
restrict our attention to small Reynolds number flows.  For groundwater
flow, Darcy's law is often used, but this does not resolve the pore
structure and assumes a homogeneous geometry which is often not valid.
Instead, we use the incompressible Stokes equations to represent the
flow, and the individual pores are resolved which results in a complex
multiply-connected geometry.  A boundary integral equation is
well-suited for solving the incompressible Stokes equations in a complex
geometry since it can resolve individual grains and the non-negligible
interactions between bodies are resolved. Moreover, high-order accuracy
and computational efficiency are obtained. To link the eroding body and
fluid dynamics, the body is eroded in the normal direction at a rate
that is proportional to the magnitude of the shear
stress~\cite{ris-moo-chi-she-zha2012}.  Tracking the interface of
fluid-structure interaction problems with a long time horizon is
particularly---the mesh is often distorted and tangled by the dynamics
and regions of high curvature develop.  These two issues are resolved
by, first, introducing a tangential velocity field to maintain a grid
that is equispaced in arclength, but that does not affect the dynamics,
and, second, a curvature penalization term.  Furthermore, the frame of
reference is changed to \thL~variables so that the curvature term
appears as a diffusive term that can easily be treated implicitly
to avoid stiffness.

@@

At the second stage of RK2, we discretize \eqref{eq:mu}--\eqref{eqn:thetaevo3} as 
\begin{align}
\label{eq:mu1d}
& \mu_k^{(n+1)} = \mu_k^{(n)} \exp \left( \eps k^2 {\Dt} \, \zeta^{(n+1/2)} \right) \\
\label{eq:L1d}
& \frac{L^{(n+1)} - L^{(n)}}{\Dt} = \Mterm^{(n+1/2)}, \\
\label{eq:th1d}
& \frac{ \mu_k^{(n+1)} \thhat_k^{(n+1)} - \mu_k^{(n)} \thhat_k^{(n)}}{\Dt} =
\mu_k^{(n+1/2)} \hat{\NLterm}_k^{(n+1/2)}.
\end{align}
Once again, we must eventually compute the geometry at step $n+1$, and so we use \eqref{eq:mu1d} to solve \eqref{eq:th1d} for $\theta^{(n+1)}$, giving
\begin{align}
\thhat_k^{(n+1)} = & \thhat_k^{(n)} \exp \left( - \epsilon k^2 \Dt \elfun^{(n+1/2)} \right) + \Dt \, \hat{\NLterm}_k^{(n+1/2)} \exp \left( - \frac{1}{2} \eps k^2 \Dt \left( 2 \elfun^{(n+1/2)} - \elfun^{(n)} \right) \right).
\end{align}
Again, both exponential terms act to reduce high-frequency oscillations in shape. We note that the above approach would generalize to higher-order Runge-Kutta methods, such as RK4.
